Reinforcement-Learning Windturbine Controller

### 2019-12-04 paper/paper.tex
Finished background section, some improvements on the qblade adapter

\usepackage{acro}
\usepackage{siunitx}
\usepackage{mathtools}
\usepackage{commath}
\input{abbreviations}

@@

As described in \cite[sec 8.3]{burtonWindEnergyHandbook2011}, the normal controller design for windturbines divides the operation of a turbine into two parts, below rated and above rated. Rated is a windspeed, in which the turbine gives maximum energy on highest aerodynamic effiency. At this point, the generator runs on its maximum torque and the blades are pitched to the optimum angle of 0 degrees. At this point, the tip-speed-ratio, which is a good indicator for loads on the blades is at maximum too.
Below that windspeed, the rotor generates less rotational torque than the maximum generator torque, thus the generator torque needs to be reduced to not slow the rotor down too much. Slowing the rotor down complicates the energy generation process, and though variable-speed wind turbines allow for some play, all windturbines are limited to a certain operational range. Fixed-speed wind turbines can only run on one single rotational speed and thus need to change the torque more agressively.
Above rated windspeed, the blades need to be pitched out to stop the rotor from spinning at faster speeds than what the turbine was designed for. The generator torque can be kept at maximum in this area, but to protect the turbine from damage, aerodynamic efficiency of the rotor is reduced by turning the blades along their long axis and thus moving their angle of attack against the incoming air into a less optimal range.

Additionally to this trivial control, optimizations to reduce vibrations and stress on the structures are implemented. The blades can be pitched individually and quickly enough to allow for different blade pitches during a single rotation of the rotor, which can be used to reduce the turbulence that hits the tower or to account for different windspeeds closer to the ground and further up in the air. Also, both generator torque and blade pitch can be used to counteract natural resonant frequencies of the structure, reducing material stress through extensive swinging of structural parts.

Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{burtonWindEnergyHandbook2011}. Below rated windspeed, the controller for torque is active, above rated the one for pitch. The parameters to those controllers can be calculated according to the laws of control theory, and the resulting controllers work reasonably close to theoretical maximum. % TODO find reference
 Still, in this work, we are trying to replace these hand-designed controllers with a reinforcement-learning algorithm.

@@

This section should aim to give a brief overview into Reinforcement Learning assumptions and derive briefly a couple of methods until in the end, we will describe \ac{DDPG}, the method we utilized to control our virtual windturbine.

@@

In every timestep, the agent supplies an action \(a\). This action could be button presses in an arcade game or blade pitch in a windturbine simulation. The mission of the agent is to pick actions which maximize cumulative reward. In other words, it should steer the environment to achieve best performance.

Additionally to the state and reward, an environment can optionally send out a done signal \(d\), which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state. In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage. As this is commonly ignored in reinforcement learning, we will also ignore it.

\subsection{Definitions}

Additionally to the terms defined by the environment, we introduce some terms which are common to reinforcement learning
\begin{itemize}
  \item \textbf{Policy} A function that maps from states to actions. There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$. If we approximate a policy by a function approximator parametrized with $\theta$, we write $\pi(s;\theta)$ or omit the parameters when clear.
  \item \textbf{Value} A value with respect to a policy and a starting state s gives the accumulative reward of following that policy from state s until infinitely in the future. $V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $. As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions. It is the expected accumulative reward of following that policy. Furthermore a value can be discounted or undiscounted. Discounting means we multiply a discounting factor $\gamma \in [0,1]$ to rewards in the fashion of $r_0 \gamma^0 + r_1 \gamma^1 + r_2 \gamma^2 ...$, which results in distant future rewards being weighted less than closer ones.
  \item \textbf{Return} If we measured a value function, that would be the return. Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions, but otherwise is equal to a value.
  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward. Though we rarely know this policy, there must always be at least one policy which gives the highest value of all possible policies in a finite \ac{MDP}
  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$. When following the optimal policy, we simply write $Q^*(s,a)$. Usually, our Q functions are discounted. Note that we can also define Values on Q-Values: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$. If we approximate Q by a function approximator parametrized with $\theta$, we write $Q(s, a;\theta)$ or omit it in case it is clear.
  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value function. These two frequently appear together, if that happens we are using an actor-critic algorithm.
\end{itemize}

@@

At first we will present a very basic algorithm for a machine learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}. It is proven, that

@@

Thus we just derived a method to train a Q function under the assumption of an optimal policy. In \cite{mnihPlayingAtariDeep2013}, the authors assumed a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space. We are presented with an infinite action space though, which is why we can't implement a greedy policy without high computational expense. So, for our work, we have to also learn a policy and not just a Q estimator. Luckily, our Q learning rule also works for other policies than the optimal policy, so let's try to derive a way to learn a policy.

@@

We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement learning algorithms, including the one we chose. The algorithm is called \ac{REINFORCE} (no joke) and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}. Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and is called \ac{PG}. \cite{degrisOffPolicyActorCritic2013} generalized that term and made an off-policy version of policy gradients. That version was then used by \cite{silverDeterministicPolicyGradient2013} and the stochastic policy was replaced by a deterministic one, the algorithm was called \ac{DPG}. This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used until finally optimizing it towards \ac{TD3} \cite{fujimotoAddressingFunctionApproximation2018}.

Let's assume we have a probablistic policy $\pi: S \rightarrow P(A)$ which is parametrized by $\theta$. An intuitive way of improving this policy could be to increase the gradient proportional to the reward an action yielded:
\begin{equation}
\theta_{i+1} = \theta_i + \alpha r(s,a) \Delta \pi_{\theta_i}(a|s)
\end{equation}
We could do this for all states we see and would end up with an iterative way of improving our policy. In fact, this is similar to the \ac{REINFORCE} algorithm (if we add our importance sampling trick from below). However, we might not know rewards for any action, state or have noisy rewards. Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step. So, with our newly gained Q, for all actions and states, do
\begin{equation}
\theta_{i+1} = \theta_i + \alpha Q(s, a)\Delta \pi_{\theta_i}(a|s)
\end{equation}
This way, actions with higher Q values will receive a higher gradient step. As Q is constant with respect to $\theta$ we don't have to derive along Q here. Problematic is though, that the Q term we used, always took the best possible actions, which could be vastly different from the actions under our policy. Thus, we need to reformulate Q to account for our policy instead of the optimal policy

@@

Now, in theory, we can not regard Q as constant with respect to $\theta$ anymore, because it is dependent on the policy. \cite{suttonPolicyGradientMethods} claim though that it can be ommitted and still regarded as constant. However there is still a problem if we integrate this new Q into our policy updates. We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment and also responsible for which experiences we see during our training. So let's assume the policy is poorly initialized at the beginning of our training and gives action 1 a probability 4 times as high as action 2, though taking action 2 would yield a 10\% better return in our Q function. Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often in our Monte-Carlo expectation draw. Though each of the gradient update steps is smaller due to the lower Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy. To correct for this oversampling bias, we could employ a trick which is called importance sampling. Because we know the bias from the probability distribution of our policy, we divide that probability from the update term:
\begin{equation}
\theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \frac{ \Delta \pi_{\theta_i}(a|s)}{\pi_{\theta_i}(a|s)}
\end{equation}
We might remember from calculus that $\frac{\Delta x}{x} = \Delta \log x$ so we simplify to
\begin{equation}
\theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \Delta \log \pi_{\theta_i}(a|s)
\end{equation}

This update equation is our basic policy gradient equation.

Before, we ignored for brevity that we are dealing with expectation calculations and that we are doing these updates under a stochastic environment, so to write it down formally correct, we need to introduce the density function of a state under a policy. According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever, and also according to them this distribution is independent of the starting state $s_0$ and only dependent on the policy and environment. \cite{silverDeterministicPolicyGradient2013} used a discounted version of this, $\rho^{\pi}(s)$, which weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$. Using the discounted version approximates the Monte-Carlo experiments we are doing better than the undiscounted version, as effectively we do not train forever on one constant policy, but change it frequently.

Having this $\rho$, let's write our objective function $J$ which we want to maximize.
\begin{equation}
J(\pi) = \int_S \rho^{\pi}(s) \int_A \pi(s, a)Q^{\pi}(s, a) dads = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[Q^{\pi}(s,a)]
\end{equation}

We explained intuitively above, and \cite{suttonPolicyGradientMethods} proves that deriving this yields
\begin{equation}
\Delta J(\pi) = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[\Delta \log \pi(a|s)Q^{\pi}(s, a)]
\end{equation}

@@

We sketched an overview in \ref{fig:ddpgsetup}. Remember that actor was another name for policy and critic another name for Q estimator. Notice how for calculating the Q targets, they used both the target actor $\mu'$ for generating actions which are then evaluated by the target critic $Q'$. This is their version of the improvement over \ac{DQN} for their instable Q function, as they also observed instabilities. Note how the Q function here does not really predict the value of following a specific policy anymore, as before we could always denote which policy Q was trained on. Also, this Q is approximated by a non-linear function approximator instead of the linear one from compatible \ac{DPG}. \cite{lillicrapCONTINUOUSCONTROLDEEP2015} do not provide a proof of convergence for this new, mixed Q update. Still, we have an off-policy way of improving the critic now. The actor is directly copied from \ac{DPG} even though it does not fulfil the \textit{compatible} property.

@@

\begin{equation}
\Delta_{\theta^{\mu}}J(\theta^{\mu}) = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, \mu(s;\theta^{\mu});\theta^Q)] = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, a;\theta^Q | a=\mu(s)) \Delta_{\theta^{\mu}}\mu(s;\theta^{\mu})]
\end{equation}

@@

Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}

@@

>>>>

\usepackage{fancyref}
\usepackage{acro}
\usepackage{siunitx}
\usepackage{mathtools}
\usepackage{commath}
\usepackage[ruled]{algorithm2e}
\input{abbreviations}

@@

\begin{figure}
  \centering
  \includegraphics{NREL-5MW.png}
  \caption{Control curve of the NREL 5MW turbine, from \cite{jonkmanDefinition5MWReference2009}}
  \label{fig:nrel5mw}
\end{figure}

As described in \cite[sec 8.3]{burtonWindEnergyHandbook2011}, the normal controller design for windturbines divides the operation of a turbine into two parts, below rated and above rated. To explain this, we added reference plot of the windturbine model which we will use in our further testing, the NREL 5MW reference turbine \cite{jonkmanDefinition5MWReference2009}. Figure \ref{fig:nrel5mw} describes several properties of that turbine in relation to windspeed. Rated windspeed is a windspeed, in which the turbine gives maximum energy on highest aerodynamic effiency. In our example figure, this point is at 11.4m/s inflowing wind. At this point, the generator runs on its maximum torque (green graph) and the blades are pitched to the optimum angle of 0 degrees (red graph). At this point, the tip-speed-ratio (purple graph) is at its designed spot. When designing the turbine blades, this tip speed ratio plays an important role.
Below that windspeed, the rotor generates less rotational torque than the maximum generator torque, thus the generator torque needs to be reduced to not slow the rotor down too much. Slowing the rotor down complicates the energy generation process, and though variable-speed wind turbines allow for some play, all windturbines are limited to a certain operational range. Fixed-speed wind turbines can only run on one single rotational speed and thus need to change the torque more aggressively. In our reference turbine, you can see how the rotor speed is kept above 6.9 rpm, which is the cut-in speed of the NREL 5MW turbine. As the generated power of a windturbine is proportional to its rotor speed times generator torque, we produce less than our rated energy in this area.
Above rated windspeed, the blades need to be pitched out to stop the rotor from spinning at faster speeds than what the turbine was designed for. See the red line in the reference plot - blades are kept at zero below rated and from then on control is done only through blade pitch change. The generator torque can be kept at maximum in this area and such will also the energy production stay at maximum. To protect the turbine from damage, aerodynamic efficiency of the rotor is reduced by turning the blades along their long axis and thus moving their angle of attack against the incoming air into a less optimal range. This is done in a way that the rotational speed of the rotor stays exactly constant.

Additionally to this trivial control, optimizations to reduce vibrations and stress on the structures are implemented. The blades can be pitched individually and quickly enough to allow for different blade pitches during a single rotation of the rotor, which can be used to reduce the turbulence that hits the tower or to account for different windspeeds closer to the ground and further up in the air. Also, both generator torque and blade pitch can be used to counteract natural resonant frequencies of the structure, reducing material stress through extensive swinging of structural parts. In our 5MW reference turbine, only blade pitch is used to counteract resonances.

Usually this control is implemented by two PID controllers, which are hand-designed as described in \cite[sec 8.4]{burtonWindEnergyHandbook2011}. PID-Controllers in general have one real-valued input over time and deliver an output. This is done by calculating an error term between the input and a desired input. The output is calculated as a mixture of the error itself (P - proportional to the error), integrating the error over the last few timesteps (I - integral of the error) and calculating a derivative to the last timestep (D - derivative). The output is the sum of these 3 terms each factorized with a constant factor.
However back to windturbines. Usually, we implement two PID controllers to control a turbine, a torque and a pitch controller. Below rated windspeed, the controller for torque is active, above rated the one for pitch. The parameters to those controllers can be calculated according to the laws of control theory. Sometimes, this strong split into two operational ranges is relaxed a bit, as in the 5MW controller, the pitch part counteracts resonances also below rated speed. From this simple concept, the resulting controllers work reasonably close to theoretical maximum already. Our reference turbine has a peak power coefficient of 0.482 - meaning that at optimal wind speed, 48\% of the energy of the wind is converted into electricity. In fact, the theoretical maximum is given by the Betz limit of 59.3\% and if we account for electrical and frictional losses, the realistic maximum is even lower. So there is not much potential to be alleviated. Also, only a fraction of this potential lies in the responsibility of the control circuit. Blade design has so far been the major area for improvements.

Still, in this work, we are trying to replace these hand-designed controllers with a reinforcement-learning algorithm. We are not expecting to exceed the performance of a industry controller. So why do we even do this?

\subsection{Why we do what we do}
As you might have seen before, all controllers follow a certain style. They take a single input variable, such as rotor speed, and adjust a single output such as torque. This type of controller is called \ac{SISO}. \ac{SISO} controllers are easy to design and work reliably. They can also be interconnected, such as with the pitch controller which gets a vibration input and a rotor speed input to combine them to one output. The main challenge with wind turbine control at the moment is however not to optimize for power, but to optimize for life-time \cite[Chapter 4]{vankuikLongtermResearchChallenges2016}. The controllers are dealing well with adjusting energy to a good level, but they are missing out on keeping the windturbine intact over a long life time.

Also, the outputs get more complex. Back in the early days of wind turbines, you could control 3 variables - pitch, torque and yaw. Nowadays, we can control each of the blades individually and with a quick enough response time to wiggle the blades around within one single rotor rotation. Research on adding active control elements to the blades such as flaps is happening, which would add many more control parameters. You might have seen flaps on an aircraft wing, extending the length of the wing or adding gaps at takeoff and landing to account for the low wind speeds hitting the aircraft wing in these flight scenarios. This could also benefit wind turbines at low wind speeds. Tackling all this with \ac{SISO} is possible, but a lot of manual work.

Next, the inputs also get more complex. It is getting cheaper to install high-precision sensors, and many of them. Wind measurement techniques such as LIDAR based aenometers are able to efficiently and precisely predict incoming wind situations. Load measurements can be done in real-time across a turbine blade and in theory, it would be possible to incorporate these into control systems. However, accomodating all these input parameters in \ac{SISO} systems is difficult as we would have to hand-model each of the inputs and how exactly it ends up in the output.

We hope to lay a foundational step in solving these issues by using an end-to-end machine learning algorithm which is capable of learning a wind turbine control scenario with many inputs and outputs. Also, we must admit that we are generally interested in playing around with reinforcement learning, it's fucking cool.

@@

So, why are we using reinforcement learning? We have posed the challenge earlier, that we want to learn a control system with a high number of inputs and outputs. Reinforcement learning has proven to be able to tackle high-complexity problems, both in input and output dimension. There have been successes in learning to play atari games based on pixel inputs \cite{mnihPlayingAtariDeep2013} and outperform humans, there have been successes in playing games with very long-term strategies such as go \cite{silverMasteringGameGo2016}, and research is continuing to improve. Because of all these successes, we try to automatically learn a windturbine controller.

Unfortunately, reinforcement learning isn't possible without a good amount of maths, so the following sections will be mainly mathematical.

@@

Additionally to the state and reward, an environment can optionally send out a done signal \(d\), which indicates a state in which further simulation is not possible and the environment wants to be reset to an initial state. In a chess game, this would be a loss or a win, and with a windturbine that could for example be fatal structural damage.

In every timestep, the agent supplies an action \(a\). This action could be button presses in an arcade game or blade pitch in a windturbine simulation. The mission of the agent is to pick actions which maximize cumulative reward. In other words, it should steer the environment to achieve best performance. Let's write down this mission. We want to maximize:

\begin{equation}
J = \mathbb{E}r
\end{equation}

This formula is still a bit incomplete - we didn't yet specify our expectation. And wasn't reward defined on states and actions? But that is because we are still missing some definitions. Luckily, there are some ahead.

\subsection{Definitions}
\label{sec:definitions}

Additionally to the terms defined by the environment, we introduce some terms which are common to reinforcement learning
\begin{itemize}
  \item \textbf{Policy} A function that maps from states to actions. There are deterministic policies $\mu: S -> A$ and stochastic policies $\pi: S -> P(A)$. If we approximate a policy by a function approximator parametrized with $\theta$, we write $\pi(s;\theta)$ or omit the parameters when clear.
  \item \textbf{Value} A value with respect to a policy and a starting state s gives the accumulative reward of following that policy from state s until infinitely in the future. $V^{\pi}(s_t) = \mathbb{E}_{s_i \sim \varepsilon, a_i \sim \pi, i>t} r(s_i, a_i) $. As we have a stochastic environment and policy, we have to add an expectation term over actions taken and states resulting from those actions. It is the expected accumulative reward of following that policy. Furthermore a value can be discounted or undiscounted. Discounting means we multiply a discounting factor $\gamma \in [0,1]$ to rewards in the fashion of $r_0 \gamma^0 + r_1 \gamma^1 + r_2 \gamma^2 ...$, which results in distant future rewards being weighted less than closer ones.
  \item \textbf{Return} If we measured a value function, that would be the return. Obviously we can't measure infinitely, so a return is only defined on a finite trajectory of state-transitions, but otherwise is equal to a value.
  \item \textbf{Optimal policy} The theoretical construct of a policy that always takes the best action with respect to accumulative reward. Though we rarely know this policy, there must always be at least one policy which gives the highest value of all possible policies in a finite \ac{MDP}
  \item \textbf{Q-Value} The reward of taking action a in state s plus the value following a policy after that $Q^{\pi}(s,a) = r(s, a) + \mathbb{E}_{s' \sim \varepsilon} V^{\pi}(s')$. When following the optimal policy, we simply write $Q^*(s,a)$. Usually, our Q functions are discounted. Note that we can also define Values on Q-Values: $V^{\pi}(s) = \mathbb{E}_{a \sim \pi} Q(s, a)$. If we approximate Q by a function approximator parametrized with $\theta$, we write $Q(s, a;\theta)$ or omit it in case it is clear.
  \item \textbf{Actor-Critic} Actor is just another name for a learnt policy and critic just another name for an estimated Q-Value function. These two frequently appear together, if that happens we are using an actor-critic algorithm.
  \item \textbf{State-density} According to \cite{suttonPolicyGradientMethods}, $d^{\pi}(s) = \lim_{t -> \infty} p(s_t = s | s_0, \pi)$ describes how likely it is to be in state s when following a policy $\pi$ forever. If we for example have a policy which plays go as well as the authors of this paper, states with a lot of enemy stones and very few own stones might be very likely, and a state without enemy stones at all would be very unlikely. Also according to \cite{suttonPolicyGradientMethods} this distribution is independent of the starting state $s_0$ and only dependent on the policy and of course the environment. In later papers, commonly a discounted version of this is used. $\rho^{\pi}(s)$ weights future probabilities less and thus can be described as the probability of being in state s \textit{soon} when following policy $\pi$. We will omit a precise definition here, you can look it up at \cite[Section 2.1]{silverDeterministicPolicyGradient2013}.
\end{itemize}

Now, with all these definitions at hand, we can specify our mission a bit better. We want to maximize
\begin{equation}
J^{\pi} = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}r(s, a) = \int_S \rho^{\pi}(s) \int_A \pi(s) r(s, a)
\label{equ:maximize_rewards}
\end{equation}

In other words, we want a very high probability of being in a state ($\rho^{\pi}$) which might give good rewards, and in that state we want a high probability of taking an action ($\pi$) which might give a good reward. In our wind turbine, rewards could for example be the all the costs and incomes summed together, so at every point in time we could count income by energy yield or costs because of maintenance (with our controller more likely costs for rebuilding the entire turbine). But before we choose $\pi$, there is a bit of a problem ahead. We can observe some rewards, but only very few of them. We might know that using 0.3MN of torque at rotational speed 11rpm is a good idea, but how about rotational speed 11.1rpm? We might have never seen that. Also, rotating the turbine at 15rpm might be a good strategy to immediately achieve high power outputs, but we might need to fix our turbine after running this policy for just a few days and incur a heavy penalty. Luckily we have defined Q-Values above. We could use them somehow, so let's discuss how to learn them.

@@

At first we will present a very basic algorithm for a Q learning agent, which is based on the basic Bellman-Equation \cite{bellmanTheoryDynamicProgramming1954}. It is proven, that

@@

Thus we just derived a method to train a Q function under the assumption of an optimal policy. In \cite{mnihPlayingAtariDeep2013}, the authors assumed a finite state space and could use the greedy policy $argmax_a Q(s', a)$ by searching the entire action space. We are presented with an infinite action space though, which is why we can't implement a greedy policy without high computational expense. So, for our work, we have to also learn a policy and not just a Q estimator. So let's try to derive a way to learn a policy.

@@

We will at first give an intuitive explanation of a basic algorithm which forms the basis for many modern reinforcement learning algorithms, including the one we chose. The algorithm is called \ac{REINFORCE} (no joke) and was originally presented by \cite{williamsSimpleStatisticalGradientfollowing}. Then we will add a Q term to it according to \cite{suttonPolicyGradientMethods} and call it \ac{PG}. \cite{degrisOffPolicyActorCritic2013} generalized that term and made an off-policy version of policy gradients, which we will not discuss in detail. That version was then used by \cite{silverDeterministicPolicyGradient2013} and the stochastic policy was replaced by a deterministic one, the algorithm was called \ac{DPG}. This algorithm finally lays the basis for \ac{DDPG} by \cite{lillicrapCONTINUOUSCONTROLDEEP2015}, which is what we used to start our experimentation.

Let's assume we have a probablistic policy $\pi: S \rightarrow P(A)$ which is parametrized by $\theta$. An intuitive way of improving this policy could be to increase the gradient proportional to the reward an action yielded:
\begin{equation}
\theta_{i+1} = \theta_i + \alpha r(s,a) \Delta \pi_{\theta_i}(a|s)
\label{equ:reinforce}
\end{equation}
We could do this for all states we see and would end up with an iterative way of improving our policy. In fact, this is similar to the \ac{REINFORCE} algorithm (if we add our importance sampling trick from below). However, we might not know rewards for any action, state or have noisy rewards. Luckily, we derived Q-Learning in the section before and thus have an estimation which action will give which return in this step. So, with our newly gained Q, for all actions and states, do
\begin{equation}
\theta_{i+1} = \theta_i + \alpha Q(s, a)\Delta \pi_{\theta_i}(a|s)
\end{equation}
This way, actions with higher Q values will receive a higher gradient step. As Q is constant with respect to $\theta$ we don't have to derive along Q here. Problematic is though, that the Q term we used, always took the best possible actions, which could be vastly different from the actions under our policy. Thus, we need to reformulate Q to account for our policy instead of the optimal policy

@@

Now, in theory, we can not regard Q as constant with respect to $\theta$ anymore, because it is dependent on the policy. \cite{suttonPolicyGradientMethods} claim though that it can be ommitted and still regarded as constant. However there is still a problem if we integrate this new Q into our policy updates. We assume that we use the same $\pi$ for exploration while doing training, so this policy is responsible for taking actions in the environment and also responsible for which experiences we see during our training. So let's assume the policy is poorly initialized at the beginning of our training and gives action 1 a probability 4 times as high as action 2, though taking action 2 would yield a 10\% better return in our Q function. Now, as we are taking action 1 4 times as often as action 2, we will also perform gradient updates on action 1 4 times as often in our Monte-Carlo expectation draw. Though each of the gradient update steps is smaller due to the lower Q factor, the higher frequency of updating action 1 will lead to action 1 getting an advantage over action 2 in this update strategy. To correct for this oversampling bias, we could employ a trick which is called importance sampling. Because we know the bias from the probability distribution of our policy, we divide that probability from the update term:
\begin{equation}
\theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \frac{ \Delta \pi_{\theta_i}(a|s)}{\pi_{\theta_i}(a|s)}
\end{equation}
We might remember from calculus that $\frac{\Delta x}{x} = \Delta \log x$ so we simplify to
\begin{equation}
\theta_{i+1} = \theta_i + \alpha Q^{\pi}(s, a) \Delta \log \pi_{\theta_i}(a|s)
\label{equ:pg_update}
\end{equation}

So, let's sum up what we have done before and formulate our goal again. Before, we ignored for brevity that we are dealing with expectation calculations and that we are doing these updates under a stochastic environment, so to write it down formally correct, we need to remember the density function of a state under a policy from section \ref{sec:definitions}. Using the discounted version approximates the Monte-Carlo experiments we are doing better than the undiscounted version, as effectively we do not train forever on one constant policy, but change it frequently, so we will use $\rho^{\pi}$ to approximate how likely we are in a state $s$ when using policy $\pi$.

Having this $\rho$, let's write our objective function $J$ which we want to maximize.
\begin{equation}
J(\pi) = \int_S \rho^{\pi}(s) \int_A \pi(s, a)Q^{\pi}(s, a) dads = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[Q^{\pi}(s,a)]
\end{equation}
This looks similar to equation \ref{equ:maximize_rewards}, just that we use Q now.

\cite{suttonPolicyGradientMethods} proves that deriving this yields
\begin{equation}
\Delta J(\pi) = \mathbb{E}_{s \sim \rho^{\pi}, a \sim \pi}[\Delta \log \pi(a|s)Q^{\pi}(s, a)]
\label{equ:pg}
\end{equation}
This equation is very famous in reinforcement learning and is called policy gradients. And by coincidence, it is equivalent to what we intuitively derived in equation \ref{equ:pg_update}. We decided to skip the actual proof and use this intuitive explanation instead, and we used the intuitive explanation in \cite[Chapter 13, Section 3, after Equation 13.8]{suttonReinforcementLearningIntroduction2018} to explain the nature of this policy gradient equation.

@@

We sketched an overview in figure \ref{fig:ddpgsetup}. Remember that actor was another name for policy and critic another name for Q estimator. Notice how for calculating the Q targets, they used both the target actor $\mu'$ for generating actions which are then evaluated by the target critic $Q'$. This is their version of the improvement over \ac{DQN} for their instable Q function, as they also observed instabilities. Note how the Q function here does not really predict the value of following a specific policy anymore, as before we could always denote which policy Q was trained on. Also, this Q is approximated by a non-linear function approximator instead of the linear one from compatible \ac{DPG}. \cite{lillicrapCONTINUOUSCONTROLDEEP2015} do not provide a proof of convergence for this new, mixed Q update. Still, we have an off-policy way of improving the critic now. The actor is directly copied from \ac{DPG} even though it does not fulfil the \textit{compatible} property.

@@

\begin{equation}
\Delta_{\theta^{\mu}}J(\theta^{\mu}) = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_{\theta^{\mu}}Q(s, \mu(s;\theta^{\mu});\theta^Q)] = \mathbb{E}_{s \sim \rho^{\beta}}[\Delta_a Q(s, a;\theta^Q | a=\mu(s)) \Delta_{\theta^{\mu}}\mu(s;\theta^{\mu})]
\end{equation}

@@

Furthermore, to aid exploration, they added noise in form of a Ornstein-Uhlenbeck process \cite{uhlenbeckTheoryBrownianMotion1930}.

For readability, we added a copy of their algorithm in Algorithm \ref{alg:vanilladdpg}, with minor adjustments to fit our notation.

\begin{algorithm}
\SetAlgoLined
Randomly initialize critic network $Q(s, a;\theta^Q)$ and actor $\mu(s;\theta^{\mu})$ with weights $\theta^Q$ and $\theta^{\mu}$\\
Initialize target networks $Q'$ and $\mu'$ with weights $\theta^{Q'} \leftarrow \theta^Q, \theta^{\mu'} \leftarrow \theta^{\mu}$\\
Initialize replay buffer $\mathcal{R}$\\

\For{episode = 1, M}{
  Initialize a random process $\mathcal{N}$ for action exploration\\
  Receive initial observation state $s_1$\\
  \For{t = 1, T}{
    Select action $a_t = \mu(s_t;\theta^{\mu}) + \mathcal{N}_t$ according to the current policy and exploration noise\\
    Execute action $a_t$ and observe reward $r_t$ and observe new state $s_{t+1}$\\
    Store transition $(s_t, a_t, r_t, s_{t+1})$ in $\mathcal{R}$\\
    Sample a random minibatch of $N$ transitions $(s_i, a_i, r_i, s_{i+1})$ from $\mathcal{R}$\\
    Set $y_i = r_i + \gamma Q'(s_{i+1}, \mu'(s_{i+1};\theta^{\mu'});\theta^{Q'})$\\
    Update critic by minimizing the loss: $L = \frac{1}{N} \sum_i (y_i - Q(s_i, a_i;\theta^Q))^2$\\
    Update the actor policy using the sampled policy gradient: $\Delta_{\theta^{\mu}} J \approx \frac{1}{N} \sum_{i} \Delta_a Q(s_i, a;\theta^Q | a=\mu(s_i)) \Delta_{\theta^{\mu}}\mu(s_i;\theta^{\mu})$\\
    Update the target networks:\\
    $\theta^{Q'} \leftarrow \tau \theta^{Q} + (1-\tau)\theta^{Q'}$\\
    $\theta^{\mu'} \leftarrow \tau \theta^{\mu} + (1-\tau)\theta^{\mu'}$
  }
}
\caption{Vanilla DDPG algorithm}
\label{alg:vanilladdpg}
\end{algorithm}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{Miniturbine.png}
  \caption{The mini wind-turbine from \cite{kolterDesignAnalysisLearning2012}}
  \label{fig:miniturbine}
\end{figure}

Their miniature model was built from cheap and low-scale parts, they simulated a wind-tunnel by attaching several fans in front of a wooden tube and at the end placed their turbine. Figure \ref{fig:miniturbine} illustrates this. Each of the blades of that miniature turbine is individually controllable and through a correction term, they are able to achieve independent pitch control which is able to pitch the motors to a value based on rotor position. Though this turbine rotates many times faster than a normal, full scale wind turbine, the underlying principles are the same (500rpm vs 11rpm). For us most importantly, they used reinforcement learning to control this turbine.

As training input, they used a single observation in contrast to popular reinforcment learning which operates on high-dimensional observation space. This observation was power output of the turbine, and they treated it as reward to the system, not having any state observations. Their action space consisted of electrical load on the generator and blade pitch, which was uniformly applied to all 3 blades. In fact, they designed their policy as to return a gaussian distribution over these two parameters directly without any input: $\pi: \varnothing \rightarrow P(\mathbb{R}^2)$. To run the training, they kept the wind speed constant, let the turbine spool up to speed and then let it run with a set of parameters returned by the policy. They ran these parameters for 2 seconds on the turbine, averaged power output over that time into a reward and then ran an iteration of their algorithm to generate new parameters, starting with random parameters. 
They first tested the \ac{REINFORCE} algorithm, with operates without a Q network and trains a policy directly on rewards. This algorithm was discussed above and utilizes the following gradient equation:

\begin{equation}
\Delta_{\pi^{\theta}} J = \mathbb{E}_{a \sim \pi, r \sim \varepsilon | a} r \Delta \log \pi(a;\theta)
\end{equation}

Note, how they denoted J in their paper as rewards, while we are using r for rewards and J for objective functions. Also, they used $\omega$ as a symbol for the action taken, while we use $a$. Also note how their policy is not dependent on a state but directly predicts actions. They did not explicitely note what they formulated their expectation term on, but we added it here for clarity. Note the similarity to what we discussed in equation \ref{equ:reinforce} and \ref{equ:pg}. In addition to this, they came up with an algorithm which they call Trust Region Policy Search, and which is astonishingly different from Trust Region Policy Optimization. With a few additions, they incorporated off-policy learning and second order error approximation into their algorithm, and showed later that it performs nicely. They were able to find optimal pitch and load settings as soon as 15 timesteps after the start of the simulation.
